Semilattice Graphs for Distributed Databases

### 2022-10-03 paper/paper.tex
cleanup

      The total space $\mathrm{Tot}(L, g)$ is a preorder whose elements consist of pairs $(x, x')$ where $x : L$ and $x' : g(x)$. We consider $(x, x') \leq (y, y')$ when $x \leq y$ and $g(x \leq y)(x') \leq y'$. The total space is a semilattice, as shown:

      \begin{theorem}
        The total space of a semilattice fibration $g$ is a semilattice.
      \end{theorem}
      \begin{proof}
        Consider two elements of the total space, $(x, x'), (y, y')$. Their join is defined as $(x, x') \vee (y, y') := (x \vee y', g(x \leq x \vee y)(x') \vee g(y \leq x \vee y)(y'))$. We must show that this join is commutative, associative, and idempotent.
        \begin{itemize}
          \item \emph{Commutative}:
            \begin{align*}
              (x, x') \vee (y, y') &= \left(x \vee y, g(x \leq x \vee y)(x') \vee g(y \leq x \vee y)(y')\right) \\
              &= \left(y \vee x, g(y \leq y \vee x)(y') \vee g(x \leq y \vee x)(x')\right) \\
              &= (y, y') \vee (x, x')
            \end{align*}
          \item \emph{Associative}:
            \begin{align*} &((x, x') \vee (y, y')) \vee (z, z')  \\
              =& \left(x \vee y \vee z, g(x \vee y \leq x \vee y \vee z)(g(x \leq x \vee y)(x') \vee g(y \leq x \vee y)(y')) \vee g(z \leq x \vee y \vee z)(z')\right) \\
              =& \left(x \vee y \vee z, g(x \leq x \vee y \vee z)(x') \vee g(y \leq x \vee y \vee z)(y') \vee g(z \leq x \vee y \vee z)(z')\right) \\
              =& \left(x \vee y \vee z, g(x \leq x \vee y \vee z)(x') \vee g(y \vee z \leq x \vee y \vee z)(g(y \leq y \vee z)(y') \vee g(z \leq y \vee z)(z'))\right) \\
              =& (x, x') \vee ((y, y') \vee (z, z'))
            \end{align*}
          \item \emph{Idempotent}: 
            \begin{align*}
              (x, x') \vee (x, x') &= \left(x \vee x, g(x \leq x \vee x)(x') \vee g(x \leq x \vee x)(x')\right) \\
              &= \left(x, g(x \leq x)(x')\right) \\
              &= (x, x')
            \end{align*}
        \end{itemize}
      \end{proof}

@@

      A semilattice fibration consists of a function $g_l : L \rightarrow \mathrm{Semilattice}$, assigning to each element of $L$ a semilattice, and a function $g_t : \Pi (x, y : L, x \leq y) g_l(x) \rightarrow g_l(y)$ (a ``transport function''), satisfying:

      \begin{itemize}
        \item For any $x : L, x' : g_l(x)$: $$g_t(x, x, x') = x'$$.
        \item For any $x, y, z : L$ such that $x \leq y \leq z$, and $x' : g_l(x)$: $$g_t(y, z, g_t(x, y, x')) = g_t(x, z, x')$$.
        \item For any $x, y : L$ such that $x \leq y$, and $x'_1, x'_2 : g_l(x)$: $$g_t(x, y, x'_1 \vee x'_2) = g_t(x, y, x'_1) \vee g_t(x, y, x'_2)$$.
      \end{itemize}

@@

      This is equivalent to the above category-theoretic definition. $g_l$ corresponds to the semilattice fibration functor's action on objects, and $g_t$ to its action on morphisms (itself a functor). The first two conditions ensure that $(g_l, g_t)$ is a functor, while the third ensures that $g_t(x, y, \cdot)$ is a semilattice morphism.

@@

      The total space $\mathrm{SemiTot}(L, g_l, g_t)$ is then a semilattice whose elements are pairs $(x, x')$ where $x : L$ and $x' : g_l(x)$. The join is defined as $(x, x') \vee (y, y') := (x \vee y, g_t(x, x \vee y, x') \vee g_t(y, x \vee y, y'))$. This is equivalent to the above category-theoretic definition, proving it is a semilattice.

@@

      \begin{align*}
        (g_{n, l}, g_{n, t}) &: \mathrm{SemiFibration}(T_{n-1}) & \text{ for natural $n \geq 1$ } \\
        T_n &:= \mathrm{SemiTot}(T_{n-1}, g_{n, l}, g_{n, t})    & \text{ for natural $n \geq 1$ }
      \end{align*}

      We now wish to define $L_\omega$ to be a semilattice that is, morally speaking, the limit of $L_n$. This lattice's elements are infinite lists $x_1, x_2, \ldots$ satisfying $x_n : g_{n, l}(x_{<n})$ for each natural $n \geq 1$, where $x_{<n} = x_1, \ldots, x_{n-1} : T_{n-1}$. The joins are defined as follows:

      $$(\overline{x} \vee \overline{y})_n := g_{n, t}(x_{<n}, x_{<n} \vee y_{<n}, x_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n})} g_{n, t}(y_{<n}, x_{<n} \vee y_{<n}, y_n)$$

@@

        To verify that $L_\omega$ is a semilattice, we must verify three properties:
        \begin{itemize}
          \item \emph{Commutativity}:
            \begin{align*}
              (\overline{x} \vee \overline{y})_n
              &= g_{n, t}(x_{<n}, x_{<n} \vee y_{<n})(x_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n})} g_{n, t}(y_{<n}, x_{<n} \vee y_{<n}, y_n) \\
              &= g_{n, t}(y_{<n}, y_{<n} \vee x_{<n})(y_n) \vee_{g_{n, l}(y_{<n} \vee x_{<n})} g_{n, t}(x_{<n}, y_{<n} \vee x_{<n}, x_n) \\
              &= (\overline{y} \vee \overline{x})_n
            \end{align*}
          \item \emph{Associativity}:
            \begin{align*}
              & ((\overline{x} \vee \overline{y}) \vee \overline{z})_n \\
              =& g_{n, t}(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, (\overline{x} \vee \overline{y})_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})}
                 g_{n, t}(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& g_{n, t}(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n},
                          g_{n, t}(x_{<n}, x_{<n} \vee y_{<n}, x_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n})}
                          g_{n, t}(y_{<n}, x_{<n} \vee y_{<n}, y_n)) \\
               &\vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} g_{n, t}(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& g_{n, t}(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, g_{n, t}(x_{<n}, x_{<n} \vee y_{<n}, x_n)) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, g_{n, t}(y_{<n}, x_{<n} \vee y_{<n}, y_n)) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& g_{n, t}(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, y_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& g_{n, t}(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, g_{n, t}(y_{<n}, y_{<n} \vee z_{<n}, y_n)) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, g_{n, t}(z_{<n}, y_{<n} \vee z_{<n}, z_n))  \\
              =& g_{n, t}(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n},
                          g_{n, t}(y_{<n}, y_{<n} \vee z_{<n}, y_n) \vee_{g_{n, l}(y_{<n} \vee z_{<n})}
                          g_{n, t}(z_{<n}, y_{<n} \vee z_{<n}, z_n)) \\
              =& g_{n, t}(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{g_{n, l}(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & g_{n, t}(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, (\overline{y} \vee \overline{z})_n) \\
              =& (\overline{x} \vee (\overline{y} \vee \overline{z}))_n
            \end{align*}
          \item \emph{Idempotence}:
            \begin{align*}
              (\overline{x} \vee \overline{x})_n &= g_{n, t}(x_{<n}, x_{<n} \vee x_{<n})(x_n) \vee_{g_{n, l}(x_{<n} \vee x_{<n}))} g_{n, t}(x_{<n}, x_{<n} \vee x_{<n}, x_n) \\
              &= g_{n, t}(x_{<n}, x_{<n}, x_n) \\
              &= x_n
            \end{align*}
        \end{itemize}
      \end{proof}

>>>>

      The total space $\mathrm{SemiTot}(L, g)$ is a preorder whose elements consist of pairs $(x, x')$ where $x : L$ and $x' : g(x)$. We consider $(x, x') \leq (y, y')$ when $x \leq y$ and $g(x \leq y)(x') \leq y'$. Consider two elements of the total space, $(x, x'), (y, y')$. Their join is defined as $$(x, x') \vee (y, y') := (x \vee y', g(x \leq x \vee y)(x') \vee g(y \leq x \vee y)(y'))$$.

      \begin{theorem}
        The total space $\mathrm{SemiTot}(L, g)$ of a semilattice $L$ and a semilattice fibration $g$ is a semilattice.
      \end{theorem}
      \begin{proof}
        We must show that this join is commutative, associative, and idempotent.
        \begin{itemize}
          \item \emph{Commutative}:
            \begin{align*}
              (x, x') \vee (y, y') &= \left(x \vee y, g(x \leq x \vee y)(x') \vee g(y \leq x \vee y)(y')\right) \\
              &= \left(y \vee x, g(y \leq y \vee x)(y') \vee g(x \leq y \vee x)(x')\right) \\
              &= (y, y') \vee (x, x')
            \end{align*}
          \item \emph{Associative}:
            \begin{align*} &((x, x') \vee (y, y')) \vee (z, z')  \\
              =& \left(x \vee y \vee z, g(x \vee y \leq x \vee y \vee z)(g(x \leq x \vee y)(x') \vee g(y \leq x \vee y)(y')) \vee g(z \leq x \vee y \vee z)(z')\right) \\
              =& \left(x \vee y \vee z, g(x \leq x \vee y \vee z)(x') \vee g(y \leq x \vee y \vee z)(y') \vee g(z \leq x \vee y \vee z)(z')\right) \\
              =& \left(x \vee y \vee z, g(x \leq x \vee y \vee z)(x') \vee g(y \vee z \leq x \vee y \vee z)(g(y \leq y \vee z)(y') \vee g(z \leq y \vee z)(z'))\right) \\
              =& (x, x') \vee ((y, y') \vee (z, z'))
            \end{align*}
          \item \emph{Idempotent}: 
            \begin{align*}
              (x, x') \vee (x, x') &= \left(x \vee x, g(x \leq x \vee x)(x') \vee g(x \leq x \vee x)(x')\right) \\
              &= \left(x, g(x \leq x)(x')\right) \\
              &= (x, x')
            \end{align*}
        \end{itemize}
      \end{proof}

@@

      A semilattice fibration consists of a function $l : L \rightarrow \mathrm{Semilattice}$, assigning to each element of $L$ a semilattice, and a function $t : \Pi (x, y : L, x \leq y) l(x) \rightarrow l(y)$ (a ``transport function''), satisfying:

      \begin{itemize}
        \item For any $x : L, x' : l(x)$: $$t(x, x, x') = x'$$.
        \item For any $x, y, z : L$ such that $x \leq y \leq z$, and $x' : l(x)$: $$t(y, z, t(x, y, x')) = t(x, z, x')$$.
        \item For any $x, y : L$ such that $x \leq y$, and $x'_1, x'_2 : l(x)$: $$t(x, y, x'_1 \vee x'_2) = t(x, y, x'_1) \vee t(x, y, x'_2)$$.
      \end{itemize}

@@

      This is equivalent to the above category-theoretic definition. $l$ corresponds to the semilattice fibration functor's action on objects, and $t$ to its action on morphisms (itself a functor). The first two conditions ensure that $(l, t)$ is a functor, while the third ensures that $t(x, y, \cdot)$ is a semilattice morphism.

@@

      The total space $\mathrm{SemiTot}(L, l, t)$ is then a semilattice whose elements are pairs $(x, x')$ where $x : L$ and $x' : l(x)$. The join is defined as $(x, x') \vee (y, y') := (x \vee y, t(x, x \vee y, x') \vee t(y, x \vee y, y'))$. This is equivalent to the above category-theoretic definition, proving it is a semilattice.

@@

      \begin{align*}
        (l_n, t_n) &: \mathrm{SemiFibration}(T_{n-1}) & \text{ for natural $n \geq 1$ } \\
        T_n &:= \mathrm{SemiTot}(T_{n-1}, l_n, t_n)    & \text{ for natural $n \geq 1$ }
      \end{align*}

      We now wish to define $L_\omega$ to be a semilattice that is, morally speaking, the limit of $L_n$. This lattice's elements are infinite lists $x_1, x_2, \ldots$ satisfying $x_n : l_n(x_{<n})$ for each natural $n \geq 1$, where $x_{<n} = x_1, \ldots, x_{n-1} : T_{n-1}$. The joins are defined as follows:

      $$(\overline{x} \vee \overline{y})_n := t_n(x_{<n}, x_{<n} \vee y_{<n}, x_n) \vee_{l_n(x_{<n} \vee y_{<n})} t_n(y_{<n}, x_{<n} \vee y_{<n}, y_n)$$

@@

        To verify that $L_\omega$ is a semilattice, we must verify three properties:
        \begin{itemize}
          \item \emph{Commutativity}:
            \begin{align*}
              (\overline{x} \vee \overline{y})_n
              &= t_n(x_{<n}, x_{<n} \vee y_{<n})(x_n) \vee_{l_n(x_{<n} \vee y_{<n})} t_n(y_{<n}, x_{<n} \vee y_{<n}, y_n) \\
              &= t_n(y_{<n}, y_{<n} \vee x_{<n})(y_n) \vee_{l_n(y_{<n} \vee x_{<n})} t_n(x_{<n}, y_{<n} \vee x_{<n}, x_n) \\
              &= (\overline{y} \vee \overline{x})_n
            \end{align*}
          \item \emph{Associativity}:
            \begin{align*}
              & ((\overline{x} \vee \overline{y}) \vee \overline{z})_n \\
              =& t_n(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, (\overline{x} \vee \overline{y})_n) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})}
                 t_n(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& t_n(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n},
                          t_n(x_{<n}, x_{<n} \vee y_{<n}, x_n) \vee_{l_n(x_{<n} \vee y_{<n})}
                          t_n(y_{<n}, x_{<n} \vee y_{<n}, y_n)) \\
               &\vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} t_n(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& t_n(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, t_n(x_{<n}, x_{<n} \vee y_{<n}, x_n)) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(x_{<n} \vee y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, t_n(y_{<n}, x_{<n} \vee y_{<n}, y_n)) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& t_n(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(y_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, y_n) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, z_n)  \\
              =& t_n(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, t_n(y_{<n}, y_{<n} \vee z_{<n}, y_n)) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, t_n(z_{<n}, y_{<n} \vee z_{<n}, z_n))  \\
              =& t_n(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n},
                          t_n(y_{<n}, y_{<n} \vee z_{<n}, y_n) \vee_{l_n(y_{<n} \vee z_{<n})}
                          t_n(z_{<n}, y_{<n} \vee z_{<n}, z_n)) \\
              =& t_n(x_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, x_n) \vee_{l_n(x_{<n} \vee y_{<n} \vee z_{<n})} \\
               & t_n(y_{<n} \vee z_{<n}, x_{<n} \vee y_{<n} \vee z_{<n}, (\overline{y} \vee \overline{z})_n) \\
              =& (\overline{x} \vee (\overline{y} \vee \overline{z}))_n
            \end{align*}
          \item \emph{Idempotence}:
            \begin{align*}
              (\overline{x} \vee \overline{x})_n &= t_n(x_{<n}, x_{<n} \vee x_{<n})(x_n) \vee_{l_n(x_{<n} \vee x_{<n}))} t_n(x_{<n}, x_{<n} \vee x_{<n}, x_n) \\
              &= t_n(x_{<n}, x_{<n}, x_n) \\
              &= x_n
            \end{align*}
        \end{itemize}
      \end{proof}
